---
title: "RNA structure"
format: html
editor_options: 
  chunk_output_type: console
---

# DROSOPHILA MELANOGASTER

* Get constitutive coordinates of Dmel transcript isoforms that will be used for genome-wide probe design.
* Use bedtools type approach on GTF file. 

> To do

* How to deal with overlapping genes on the same strand?

## Environment

```{r}
library(tidyverse)
library(valr)
library(furrr)
library(qs)
library(patchwork)

plan(multisession, workers = 8)

release_version <- 113

```

## Process GTF

### Get GTF and prepare an annotated list per protein-coding and non-coding genes

* Using the latest ENSEMBL **release 113** version
* ENSEMBL GTF is 1-based (use default read_gtf() option to convert automatically to 0-based coordinates)

```{r}
## Download gtf
release_version <- 113
gtf_file <- biomartr::getGTF(
  db = "ensembl",
  organism = "Drosophila melanogaster",
  path = tempdir(),
  release = release_version
)

## import gtf
gtf <- read_gtf(gtf_file) %>%
  filter(str_detect(gene_id, "FBgn"))
head(gtf)

## prepare tx2gene file
t2g <- gtf %>%
  dplyr::select(gene_id, transcript_id) %>%
  filter(!is.na(gene_id)) %>%
  filter(!is.na(transcript_id)) %>%
  distinct()

write_csv(t2g, paste0("./GENOMEDIR/Dmel_ensembl_v", release_version,"_tx2gene.csv"))

## filter for protein coding and ncRNA
gtf_pcnc <- gtf %>% 
  filter(gene_biotype %in% c("protein_coding", "ncRNA"))

## how many genes have both +/- strand attributes? - FBgn0002781 mod(mdg4) is the only one. Ignore this outlier exon.
## This is not present in the latest v113 version
strand_test_df <- gtf_pcnc %>%
  filter(type == "exon") %>%
  group_by(gene_id) %>%
  summarise(strand = unique(strand) %>% sort() %>% paste0(collapse = ""))

gtf_pcnc <- gtf_pcnc %>%
  filter(!(gene_id == "FBgn0002781" & strand == "+"))

## create a list for each gene
pcnc_list <- gtf_pcnc %>%
  split(gtf_pcnc$gene_id)

## create a filtered list of exons
pcnc_exon_list <- pcnc_list %>%
  future_imap(~{
    exon_intervals <- .x %>% filter(type == "exon")
    isoform_count <- unique(exon_intervals$transcript_id) %>% length()
    strand <- unique(exon_intervals$strand) %>% unique()
    
    output <- list(
      gene_id = .y,
      gene_name = unique(exon_intervals$gene_name),
      gene_biotype = unique(exon_intervals$gene_biotype),
      exon_intervals = exon_intervals,
      isoform_count = isoform_count,
      strand = strand
    )
    
    return(output)
  }, .progress = TRUE)

## save the list 
qsave(pcnc_exon_list, paste0("./data/RNA-structure/Dmel/Dmel_ensembl_v",
                             release_version,
                             "_proteincoding_noncoding_exon_list.qs")
)

## distribution of transcript isoform count per gene
pcnc_exon_list <- qread("./data/RNA-structure/Dmel/Dmel_ensembl_v113_proteincoding_noncoding_exon_list.qs")

isoform_count_df <- pcnc_exon_list %>%
  future_imap_dfr(~{
    tibble(
      gene_name = pluck(.x, "gene_name"),
      gene_biotype = pluck(.x, "gene_biotype"),
      isoform_count = pluck(.x, "isoform_count")
      )
  }, .id = "gene_id")

if(!dir.exists("./data/RNA-structure/Dmel/plots")){
  dir.create("./data/RNA-structure/Dmel/plots")
}

isoform_count_df %>%
  ggplot(aes(x = isoform_count)) + 
  geom_histogram(bins = 50) + 
  scale_y_log10() + 
  labs(title = "Number of tx isoforms per gene") + 
  facet_wrap(~ gene_biotype) + 
  theme_bw(base_size = 15)

ggsave("./data/RNA-structure/Dmel/plots/plot_tx-isoform-count-per-gene.pdf")

```

### Find constitutive intervals for multi-isoform transcripts

#### set up bedtools environment

* My conda env has bedtools 2.31.1, whereas bedtoolsr is built for 2.30.0 - just in case there are syntax errors...

```{r}
library(bedtoolsr)

conda_bedtools_bin <- "/opt/miniconda3/envs/bedtools/bin" # find this directory
Sys.setenv(PATH = paste(Sys.getenv("PATH"), conda_bedtools_bin, sep = .Platform$path.sep))
Sys.getenv("PATH")
Sys.which("bedtools")
```

#### find multi-intersecting regions 

```{r}
##
pcnc_exon_list <- qread("./data/RNA-structure/Dmel/Dmel_ensembl_v113_proteincoding_noncoding_exon_list.qs") 

## find constitutive regions and create a list
pcnc_constitutive_list <- pcnc_exon_list %>%
  map(~{
    if(.x$isoform_count > 1){
      ## multiinter output
      multiinter_output <- pluck(.x, "exon_intervals") %>%
        split(.$transcript_id) %>%
        map(valr::bed_sort) %>%
        bedtoolsr::bt.multiinter(header = TRUE, names = paste(names(.), collapse = " ")) %>% 
        as_tibble() %>%
        janitor::row_to_names(row_number = 1) %>%
        mutate(across(c(start, end, num), as.numeric)) %>%
        mutate(strand = pluck(.x, "strand")) %>%
        mutate(width = end - start) %>%
        dplyr::select(chrom, start, end, width, strand, everything())
      
      .x$multiinter_output <- multiinter_output
      
      ## constitutive intervals
      constitutive_intervals <- multiinter_output %>%
        filter(num == pluck(.x, "isoform_count")) %>%
        dplyr::select(chrom, start, end, width, strand)
      
      .x$constitutive_intervals <- constitutive_intervals
      
      ## does constitutive intervals exist
      .x$constitutive_exist <- ifelse(nrow(constitutive_intervals) > 0, TRUE, FALSE)
      
      ## length of constitutive intervals
      .x$constitutive_length <- ifelse(
        nrow(constitutive_intervals) == 0,
        0,
        sum(constitutive_intervals$width)
      )
      
      return(.x)
      
    } else if(.x$isoform_count == 1){
      .x$multiinter_output <- pluck(.x, "exon_intervals") %>%
        mutate(num = 1) %>%
        dplyr::select(chrom, start, end, width, strand, num, "list" = transcript_id)
      
      .x$constitutive_intervals <- pluck(.x, "exon_intervals") %>%
        dplyr::select(chrom, start, end, width, strand)
      
      .x$constitutive_exist <- TRUE
      
      .x$constitutive_length <- pluck(.x, "exon_intervals") %>%
        pull(width) %>% sum()
      
      return(.x)
      
    }
  }, .progress = TRUE)

## save output
qsave(pcnc_constitutive_list, paste0("./data/RNA-structure/Dmel/Dmel_ensembl_v",
                                     release_version,
                                     "_proteincoding_noncoding_constitutive-region_list.qs")
)

```


## Extract RNA sequences 

### Download FASTA file

```{r}
release_version

fasta_file <- biomartr::getGenome(
  db = "ensembl",
  organism = "Drosophila melanogaster",
  release = release_version,
  gunzip = TRUE,
  path = tempdir(),
)

```

### Get transcript sequences of constitutive regions

```{r}
## constitutive list
pcnc_constitutive_list <- qread("./data/RNA-structure/Dmel/Dmel_ensembl_v",
                                release_version,
                                "_proteincoding_noncoding_constitutive-region_list.qs")

# bed_example <- pcnc_constitutive_list$"FBgn0025790"$constitutive_intervals %>%
#   dplyr::mutate(name = str_c(chrom, start, end, strand, sep = ":")) %>%
#   dplyr::select(chrom, start, end, name, "score" = width, strand)

# bt.getfasta(
#   fi = "~/Documents/Data/GENOMEDIR/Dmel/Drosophila_melanogaster.BDGP6.28.dna.toplevel.fa",
#   bed = pcnc_constitutive_list$"FBgn0025790"$constitutive_intervals,
#   bedOut = TRUE,
#   s = NULL
# )

# bt.getfasta(
#   fi = "~/Documents/Data/GENOMEDIR/Dmel/Drosophila_melanogaster.BDGP6.28.dna.toplevel.fa",
#   bed = bed_example,
#   bedOut = TRUE,
#   s = TRUE
# )

pcnc_constitutive_list[[312]]

## get fasta sequence, arrange order depending on the strand, and add character counts with "N"
pcnc_constitutive_seq_list <- pcnc_constitutive_list %>%
  map(~ {
    if (.x$constitutive_exist == TRUE) {
      bed_touse <- .x$constitutive_intervals %>%
        dplyr::mutate(name = str_c(chrom, start, end, strand, sep = ":")) %>%
        dplyr::select(chrom, start, end, name, "score" = width, strand)

      getfasta_output <- bedtoolsr::bt.getfasta(
        fi = fasta_file,
        bed = bed_touse,
        bedOut = TRUE,
        s = TRUE
      ) %>%
        as_tibble() %>%
        setNames(c("chrom", "start", "end", "name", "width", "strand", "seq")) %>%
        dplyr::select(-name) %>%
        dplyr::arrange(if (.x$strand == "+") start else desc(start)) %>%
        mutate(nchar_plus_N_cumsum = cumsum(width + 1))

      .x$constitutive_sequence_df <- getfasta_output
      return(.x)
    } else {
      .x$constitutive_sequence_df <- tibble(
        chrom = character(),
        start = numeric(),
        end = numeric(),
        width = numeric(),
        strand = character(),
        seq = character(),
        nchar_plus_N_cumsum = numeric()
      )
      return(.x)
    }
  }, .progress = TRUE)

## get concatenated target sequence collapsed for "N"
pcnc_constitutive_seq_list <- pcnc_constitutive_seq_list %>%
  map(~ {
    if (.x$constitutive_exist == TRUE) {
      constitutive_seq_Ncollapsed <- .x$constitutive_sequence_df %>%
        pull(seq) %>%
        paste0(collapse = "N")
      .x$constitutive_seq_Ncollapsed <- constitutive_seq_Ncollapsed
      return(.x)
    } else {
      .x$constitutive_seq_Ncollapsed <- NULL
      return(.x)
    }
  }, .progress = TRUE)

pcnc_constitutive_seq_list[[300]]

## save 
pcnc_constitutive_seq_list %>%
  qsave(paste0("./data/RNA-structure/Dmel/Dmel_ensembl_v",
               release_version,
               "_proteincoding_noncoding_constitutive-sequence_list.qs")
  )

```

## Exploring the constitutive region dataset

### Summary of isoform constitutive mappings

```{r}
pcnc_constitutive_seq_list <- 
  qread(paste0("./data/RNA-structure/Dmel/Dmel_ensembl_v",
               release_version,
               "_proteincoding_noncoding_constitutive-sequence_list.qs")
  )

## Summary dataframe
pcnc_constitutive_summary_df <- pcnc_constitutive_seq_list %>%
  imap_dfr(~{
    tibble(
      gene_id = .x$gene_id,
      gene_name = .x$gene_name,
      gene_biotype = .x$gene_biotype,
      strand = .x$strand,
      isoform_count = .x$isoform_count,
      constitutive_exist = .x$constitutive_exist,
      constitutive_length = .x$constitutive_length,
      constitutive_seq_Ncollapsed = ifelse(is.null(.x$constitutive_seq_Ncollapsed), 
                                           "NA", 
                                           as.character(.x$constitutive_seq_Ncollapsed))
    )
  }, .progress = TRUE) %>%
  mutate(constitutive_exist_over_500 = if_else(constitutive_length > 500, TRUE, FALSE)) %>%
  mutate(constitutive_exist_over_2k = if_else(constitutive_length > 2000, TRUE, FALSE))

write_csv(pcnc_constitutive_summary_df, paste0("./data/RNA-structure/Dmel/Dmel_ensembl_v",
                                               release_version,
                                               "_proteincoding_noncoding_constitutive-sequence_summary.csv")
)

## Summary plots
pcnc_constitutive_summary_df %>%
  ggplot(aes(x = gene_biotype, fill = constitutive_exist_over_500)) +
  geom_bar(position = position_stack()) + 
  geom_text(stat = "count", aes(label = after_stat(count)), position = position_stack(), vjust = -0.5) + 
  labs(title = "Constitutive sequence over 500nt exist?") + 
  theme_bw(base_size = 15) 

ggsave("./data/RNA-structure/Dmel/plots/plot_constitutive-length-over-500nt.pdf")

pcnc_constitutive_summary_df %>%
  ggplot(aes(x = gene_biotype, fill = constitutive_exist_over_2k)) +
  geom_bar(position = position_stack()) + 
  geom_text(stat = "count", aes(label = after_stat(count)), position = position_stack(), vjust = -0.5) + 
  labs(title = "Constitutive sequence over 2000nt exist?") + 
  theme_bw(base_size = 15) 

ggsave("./data/RNA-structure/Dmel/plots/plot_constitutive-length-over-2000nt.pdf")

pcnc_constitutive_summary_df %>%
  ggplot(aes(x = constitutive_length)) +
  geom_histogram() + 
  scale_x_log10() +
  facet_wrap(~ constitutive_exist + gene_biotype) +
  labs(title = "Have constitutive sequence?") + 
  theme_bw(base_size = 15)

ggsave("./data/RNA-structure/Dmel/plots/plot_constitutive-length-histogram.pdf")

pcnc_constitutive_summary_df %>% summary()

```

### Plotting transcript architecture

```{r}
pcnc_constitutive_seq_list <- 
  qread(paste0("./data/RNA-structure/Dmel/Dmel_ensembl_v",
               release_version,
               "_proteincoding_noncoding_constitutive-sequence_list.qs")
  )

## helper function
plot_isoforms <- function(list_element){
  list_element$exon_intervals %>% ggplot() +
      geom_segment(
        aes(x = min(list_element$exon_intervals$start), 
            xend = max(list_element$exon_intervals$end), 
            y = transcript_id, 
            yend = transcript_id),
        linewidth = 0.5, colour = "gray80"
      ) +
      geom_segment(
        data = list_element$exon_intervals,
        aes(x = start, xend = end, y = transcript_id, yend = transcript_id),
        linewidth = 3, colour = "gray30",
        inherit.aes = FALSE
      ) +
      geom_segment(
        data = list_element$constitutive_intervals, 
        aes(x = start, xend = end, y = "CONSTITUTIVE", yend = "CONSTITUTIVE"),
        linewidth = 7, colour = "coral", 
        inherit.aes = FALSE
      ) + 
      labs(
        title = paste0(
          list_element$gene_id, " ", 
          list_element$gene_name, " ", 
          list_element$strand, "ve strand ",
          paste0("(constitutive ", list_element$constitutive_length, "nt)")
        ),
        x = "Chromosomal coordinate",
        y = "Transcript ID"
      ) + 
      theme_bw()
}

## Explore some plots
library(patchwork)

pcnc_constitutive_seq_list %>%
  purrr::keep(~ .x$gene_name %in% c("pros", "Syp", "Pdf", "dlg1")) %>%
  map(plot_isoforms) %>%
  wrap_plots()

ggsave("./data/RNA-structure/Dmel/plots/plot_constitutive_txplot-example.pdf", width = 12, height = 9)

pcnc_constitutive_seq_list %>%
  purrr::keep(~.x$gene_name == "Myc") %>%
  pluck(1) %>%
  plot_isoforms()
  
```

### Retrieving constitutive sequence of a gene 

```{r}
pcnc_constitutive_seq_list <- 
  qread(paste0("./data/RNA-structure/Dmel/Dmel_ensembl_v",
               release_version,
               "_proteincoding_noncoding_constitutive-sequence_list.qs")
  )

gene_id <- "FBgn0262656"
output_fasta_path <- "~/Desktop/Myc.fa"

fasta_lines <- c(
  paste(
   ">", gene_id, 
   pluck(pcnc_constitutive_seq_list, gene_id) %>% pluck("gene_name"),  
   "RNA isoform-constitutive sequence (N collapsed)" 
  ),
  pluck(pcnc_constitutive_seq_list, gene_id) %>% pluck("constitutive_seq_Ncollapsed")
)

write_lines(fasta_lines, output_fasta_path)

```


## Find candidate genes for initial testing

### Brain expression

#### samuels 2020 L3 brain RNA-seq (ensembl 99)

```{r}
# list.files("~/Documents/Data/sequencing/RNA-seq/samuels2020/")
# 
# raw <- list.dirs("~/Documents/Data/sequencing/RNA-seq/samuels2020/", recursive = FALSE) %>%
#   set_names() %>%
#   map(~{
#     list.files(.x, pattern = "*.tsv", full.names = TRUE)
#   }) %>%
#   map_dfr(~{
#     read_tsv(.x) %>%
#       dplyr::select(target_id, tpm)
#     }) %>%
#   group_by(target_id) %>%
#   summarise(tpm = mean(tpm, na.rm = TRUE)) %>%
#   ungroup() %>%
#   mutate(target_id = str_replace(target_id, "-I\\.[:digit:]+", "")) %>%
#   group_by(target_id) %>%
#   summarise(tpm = sum(tpm))
# 
# ens99_t2g <- read_csv("~/Documents/LCLIP/GENOMEDIR/Dmel_tx2gene_ENSEMBL_v99.csv")
# 
# l3brain_tpm <- ens99_t2g %>%
#   left_join(raw, by = c("tx_id" = "target_id")) %>%
#   dplyr::select(-tx_id) %>%
#   group_by(gene_id, gene_name, gene_biotype) %>%
#   summarise(tpm = sum(tpm))
# 
# l3brain_tpm %>%
#   write_csv("~/Documents/Data/sequencing/RNA-seq/samuels2020/samuels2020_l3brain_orr_rna-seq_tpm.csv")

```


#### dillon 2022 scRNA-seq (ensembl 99)

```{r}
library(Seurat)
dillon_atlas <- readRDS("~/Documents/Data/sequencing/scRNA-seq/dillon2022/RDS_files/atlas_combined_annotated.rds")

dillon_progenitors_clusters_vector <- c(7,11,32,33,42,6,14,1,40,41,20,24,16,17,47,15)
dillon_glia_clusters_vector <- c('9','19','25','26','29','36','37','44') %>% as.double()
dillon_neurons_clusters_vector <- c(0,2,3,4,5,8,10,12,13,18,21,22,23,28,27,30,34,35,39,46,48,50,51,52,53)

dillon_atlas@meta.data <- dillon_atlas@meta.data %>%
  mutate(broad_celltype = case_when(
    seurat_clusters %in%dillon_progenitors_clusters_vector ~ "progentors",
    seurat_clusters %in%dillon_glia_clusters_vector ~ "glia",
    seurat_clusters %in%dillon_neurons_clusters_vector ~ "neurons",
    TRUE ~ "NA"
  )
  )

dillon_atlas <- subset(dillon_atlas, broad_celltype != "NA")
dillon_atlas@meta.data$orig.ident <- dillon_atlas@meta.data$broad_celltype
dillon_atlas %>% DimPlot(group.by = "orig.ident", reduction = "umap")

dillon_atlas <- SetIdent(dillon_atlas, value = dillon_atlas@meta.data$broad_celltype)
dillon_atlas@active.ident

## find markers
broad_markers <- FindAllMarkers(
  object = dillon_atlas, 
  assay = "RNA", 
  logfc.threshold = 0.1, 
  test.use = "wilcox",
  min.pct = 0.1, 
  min.diff.pct = 0.1, 
  verbose = TRUE, 
  return.thresh = 0.0501
)

write_csv(broad_markers, "./data/RNA-structure/Dmel/tissue_expression/dillon2022_larval-brain_all-markers.csv")

## curate best markers
brain_markers_curated <- broad_markers %>%
  as_tibble() %>%
  dplyr::select("gene_name" = gene, everything()) %>%
  filter(avg_log2FC > 1 & p_val_adj < 0.01) %>%
  group_by(cluster) %>%
  dplyr::slice_max(order_by = avg_log2FC, n = 100) %>% ungroup()

nonoverlapping_markers <- brain_markers_curated %>%
  group_by(gene_name) %>%
  summarise(n = n()) %>%
  arrange(desc(n)) %>%
  filter(n == 1) %>% pull(gene_name)

brain_markers_curated <- brain_markers_curated %>%
  filter(gene_name %in% nonoverlapping_markers) %>%
  dplyr::select(gene_name, "brain_marker" = cluster)

##
brain_tpm <- read_csv("~/Documents/Data/sequencing/RNA-seq/samuels2020/samuels2020_l3brain_orr_rna-seq_tpm.csv")

brain_tpm_marker <- brain_tpm %>%
  ungroup() %>%
  left_join(brain_markers_curated) %>%
  dplyr::select(-gene_biotype)

brain_tpm_marker %>% write_csv("./data/RNA-structure/Dmel/tissue_expression/larval_brain_tpm_marker.csv")

```

## Larval blood expression

### Cho 2020 scRNA-seq

* PH - prohemocytes
* PM - plasmatocytes
* LM - lamellocytes
* CC - crystal cells

#### find markers 

```{r}
library(Seurat)

## only retain 4 major cell types
obj <- readRDS("~/Documents/Data/sequencing/scRNA-seq/cho2020/allnew20210215_circulation.combined.indep_harmony.Rds")
obj <- UpdateSeuratObject(obj)

obj@meta.data$orig.ident <- obj@meta.data$celltype
obj %>% DimPlot(group.by = "orig.ident", reduction = "tsne")

obj <- SetIdent(obj, value = obj@meta.data$celltype)
obj@active.ident

celltype_interest <- c("PH", "PM", "LM", "CC") 
obj_subset <- subset(obj, celltype %in% celltype_interest)

obj_subset %>% DimPlot(group.by = "orig.ident", reduction = "tsne", label = TRUE, label.color = "black", label.box = TRUE)

## find markers
obj_all_markers <- FindAllMarkers(
  object = obj_subset, 
  assay = "RNA", 
  logfc.threshold = 0.1, 
  test.use = "wilcox",
  min.pct = 0.1, 
  min.diff.pct = 0.1, 
  verbose = TRUE, 
  return.thresh = 0.0501
)

write_csv(obj_all_markers, "./data/RNA-structure/Dmel/tissue_expression/cho2020_larval-blood_all-markers.csv")

## curate best markers
blood_markers_curated <- obj_all_markers %>%
  as_tibble() %>%
  dplyr::select("gene_name" = gene, everything()) %>%
  filter(avg_log2FC > 1 & p_val_adj < 0.01) %>%
  group_by(cluster) %>%
  dplyr::slice_max(order_by = avg_log2FC, n = 100) %>% ungroup()

nonoverlapping_markers <- blood_markers_curated %>%
  group_by(gene_name) %>%
  summarise(n = n()) %>%
  arrange(desc(n)) %>%
  filter(n == 1) %>% pull(gene_name)

blood_markers_curated <- blood_markers_curated %>%
  filter(gene_name %in% nonoverlapping_markers) %>%
  dplyr::select(gene_name, "blood_marker" = cluster)

```

#### pseudobulk expression levels

```{r}
## get counts per million (cpm) df
aggregate_expr <-  AggregateExpression(
  obj_subset,
  normalization.method = "RC",
  scale.factor = 1
) 

cpm_df <- aggregate_expr$RNA %>% 
  as.data.frame() %>%
  rownames_to_column(var = "gene_name") %>%
  mutate(count_sum = rowSums(across(where(is.numeric)))) %>%
  dplyr::select(gene_name, count_sum) %>%
  mutate(cpm = count_sum / sum(count_sum) * 1e6) %>%
  as_tibble()

## combine with markers
blood_cpm_marker <- cpm_df %>%
  dplyr::select(-count_sum) %>%
  left_join(blood_markers_curated)

blood_cpm_marker %>% write_csv("./data/RNA-structure/Dmel/tissue_expression/larval_blood_cpm_marker.csv")

```


### Find candidate genes based on brain/blood

#### make a df with rna structure and brain/blood expression information

```{r}
##
pcnc_constitutive_summary_df <- read_csv("./data/RNA-structure/Dmel/Dmel_ensembl_v113_proteincoding_noncoding_constitutive-sequence_summary.csv")

##
pcnc_constitutive_summary_df %>%
  arrange(desc(constitutive_length)) %>%
  slice_head(n = 10) %>%
  pull(gene_name) 

## add brain/blood dataset
pcnc_constitutive_brain_blood <- pcnc_constitutive_summary_df %>%
  left_join(brain_tpm_marker %>% dplyr::select(gene_id, "brain_tpm" = tpm, brain_marker)) %>%
  left_join(blood_cpm_marker %>% dplyr::rename("blood_cpm" = cpm))

pcnc_constitutive_brain_blood %>% write_csv("./data/RNA-structure/Dmel/small_scale_trial/pcnc_constitutive_brain_blood.csv")
```

#### selection purely on expression & length

```{r}
pcnc_constitutive_brain_blood <- read_csv("./data/RNA-structure/Dmel/small_scale_trial/pcnc_constitutive_brain_blood.csv")

## does expression correlate btwn blood and brain?
pcnc_constitutive_brain_blood %>%
  filter(brain_tpm > 1e0 & blood_cpm > 1e0) %>% 
  ggplot(aes(x = brain_tpm, y = blood_cpm)) + 
  geom_point(
    stroke = 0, alpha = 0.8, size = 1.2) + 
  ggpubr::stat_cor(show.legend = FALSE) +
  scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) + 
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) + 
  labs(
    title = "Brain vs Blood expression correlation per gene",
    x = "L3 brain (TPM, Transcripts per million)",
    y = "L3 blood (CPM, Counts per million)"
  ) +
  theme_bw(base_size = 15)

## does expression correlate with constitutive length?
pcnc_constitutive_brain_blood %>%
  ggplot(aes(x = brain_tpm, y = constitutive_length)) + 
  geom_point(stroke = 0, alpha = 0.3) + 
  ggpubr::stat_cor() +
  scale_x_log10() + 
  scale_y_log10() + 
  labs(
    title = "Brain expression vs constitutive length per gene",
    x = "L3 brain (TPM)",
    y = "Constitutive seq length (nt) log10"
  ) +
  theme_bw()

pcnc_constitutive_brain_blood %>%
  ggplot(aes(x = blood_cpm, y = constitutive_length)) + 
  geom_point(stroke = 0, alpha = 0.3) + 
  ggpubr::stat_cor() +
  scale_x_log10() + 
  scale_y_log10() + 
  labs(
    title = "Blood expression vs constitutive length per gene",
    x = "L3 blood (CPM)",
    y = "Constitutive seq length (nt) log10"
  ) +
  theme_bw()

## How many genes can we recover?
brain_tpm_min <- 40
brain_tpm_max <- 500

blood_cpm_min <- 50
blood_cpm_max <- 1500

constitutive_length_cutoff <- 2000

pcnc_constitutive_brain_blood_expr_filt <- pcnc_constitutive_brain_blood %>%
  filter(
    brain_tpm > brain_tpm_min &
    brain_tpm < brain_tpm_max &
    blood_cpm > blood_cpm_min &
    blood_cpm < blood_cpm_max 
  ) %>%
  filter(constitutive_length > constitutive_length_cutoff) %>%
  arrange(desc(constitutive_length))
pcnc_constitutive_brain_blood_expr_filt %>%
  dplyr::select(-constitutive_seq_Ncollapsed) %>% 
write_csv( "~/Desktop/pcnc_constitutive-isoform_blood_brain_filtered.csv")

pcnc_constitutive_brain_blood_expr_filt %>%
  {ggplot(data = ., aes(constitutive_length)) + 
  geom_histogram() + labs(title = paste0(as.character(nrow(.)), " genes"))}

pcnc_constitutive_brain_blood_expr_filt %>%
  ggplot(aes(x = "", y = constitutive_length, label = gene_name, colour = blood_marker)) +
  ggrepel::geom_text_repel(max.overlaps = 50, size = 6) +
  facet_wrap(~blood_marker, nrow = 1) +
  theme_bw(base_size = 15)

pcnc_constitutive_brain_blood %>%
  filter(brain_tpm > 1e0 & blood_cpm > 1e0) %>% 
  ggplot(aes(x = brain_tpm, y = blood_cpm)) + 
  geom_point(aes(colour = log10(constitutive_length)),
    stroke = 0, alpha = 0.8, size = 1.2) + 
  # ggpubr::stat_cor(show.legend = FALSE) +
  geom_hline(yintercept = c(50, 1500)) + 
  geom_vline(xintercept = c(40, 500)) +
  scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) + 
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) + 
  labs(
    title = "",
    x = "L3 brain (TPM, Transcripts per million)",
    y = "L3 blood (CPM, Counts per million)"
  ) +
  colorspace::scale_colour_continuous_sequential(palette = "inferno", rev = FALSE) +
  # facet_wrap(~ constitutive_exist_over_2k) + 
  theme_bw(base_size = 15) + 
  theme(
    legend.position = "none"
  )

```

#### write broken down fasta per 1 kb

* for testing with stellaris

```{r}
##
pcnc_constitutive_summary_df <- read_csv("./data/RNA-structure/Dmel/Dmel_ensembl_v113_proteincoding_noncoding_constitutive-sequence_summary.csv")

## function
write_fasta_chunks <- function(gene_name, size, output_fasta_path){
  constitutive_seq_Ncollapsed <- pcnc_constitutive_summary_df %>%
    filter(!!gene_name == gene_name) %>%
    pull(constitutive_seq_Ncollapsed)
  
  char_count <- nchar(constitutive_seq_Ncollapsed)
  
  # seqs <- str_extract_all(constitutive_seq_Ncollapsed, pattern = paste0('.{1,', size, '}'))
  if((char_count / size)%%1==0){
    seq1 <- seq(1, char_count, size)
    seq2 <- seq(size, char_count, size)
  } else {
    seq1 <- seq(1, char_count, size)
    seq2 <- c(seq(size, char_count, size), char_count)
  }
  
  seqs <- substring(constitutive_seq_Ncollapsed, seq1, seq2)
  
  fasta_header_intermediate <- c(1: length(seqs)) 
  fasta_headers <- paste(
    ">", gene_name, fasta_header_intermediate, "of", length(fasta_header_intermediate), 
    "(", size, "nt chuncks )",
    "constitutive N collapsed seq"
  )
  
  fasta_lines <- c()
  for(i in 1:length(seqs)){
    fasta_lines <- c(
      fasta_lines, 
      fasta_headers[i],
      seqs[i],
      ""
    )
  }
  
  write_lines(fasta_lines, output_fasta_path)
}

write_fasta_chunks("stg", 1000, "~/Desktop/stg.fa")

## does expression correlate btwn blood and brain?
pcnc_constitutive_brain_blood %>%
  filter(brain_tpm > 1e0 & blood_cpm > 1e0) %>% 
  ggplot(aes(x = brain_tpm, y = blood_cpm)) + 
  geom_point(aes(colour = log10(constitutive_length)),
    stroke = 0, alpha = 0.8, size = 1.2) + 
  ggpubr::stat_cor(show.legend = FALSE) +
  geom_hline(yintercept = c(50, 1500)) + 
  geom_vline(xintercept = c(40, 500)) +
  scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) + 
  scale_y_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) + 
  labs(
    title = "Brain vs Blood expression correlation per gene",
    x = "L3 brain (TPM, Transcripts per million)",
    y = "L3 blood (CPM, Counts per million)"
  ) +
  colorspace::scale_colour_continuous_sequential(palette = "inferno", rev = FALSE) +
  # facet_wrap(~ constitutive_exist_over_2k) + 
  theme_bw(base_size = 15)

```


#### selection on marker expression & length 

```{r}
##
pcnc_constitutive_brain_blood %>%
  ggplot(aes(y = constitutive_length)) +
  ggbeeswarm::geom_quasirandom(
    aes(x = brain_marker)
  ) + 
  ggbeeswarm::geom_quasirandom(
    aes(x = blood_marker)
  )


pcnc_constitutive_brain_blood %>%
  filter(brain_tpm > 0) %>%
  filter(!is.na(brain_marker)) %>%
  ggplot(aes(x = brain_tpm, y = constitutive_length, colour = brain_marker)) + 
  geom_point(stroke = 0, alpha = 1, size = 1) + 
  geom_text(aes(label = gene_name)) + 
  scale_x_log10() + 
  scale_y_log10() + 
  # labs(
  #   title = "Brain vs Blood expression correlation per gene",
  #   x = "L3 brain (TPM)",
  #   y = "L3 blood (CPM)"
  # ) +
  theme_bw()

pcnc_constitutive_brain_blood %>%
  filter(blood_cpm > 0) %>%
  filter(!is.na(blood_marker)) %>%
  ggplot(aes(x = blood_cpm, y = constitutive_length, colour = blood_marker)) + 
  geom_point(stroke = 0, alpha = 1, size = 1) + 
  geom_text(aes(label = gene_name)) + 
  scale_x_log10() + 
  scale_y_log10() + 
  # labs(
  #   title = "Brain vs Blood expression correlation per gene",
  #   x = "L3 brain (TPM)",
  #   y = "L3 blood (CPM)"
  # ) +
  theme_bw()



```


\newpage

# MUS MUSCULUS

## Environment

```{r}
library(tidyverse)
library(valr)
library(furrr)
library(qs)
library(patchwork)

plan(multisession, workers = 8)

release_version <- 113

```

## Process GTF

### Get GTF and prepare an annotated list per protein-coding and non-coding genes

* Using the latest ENSEMBL **release 113** version
* ENSEMBL GTF is 1-based (use default read_gtf() option to convert automatically to 0-based coordinates)

```{r}
## Download gtf
release_version <- 113
gtf_file <- biomartr::getGTF(
  db = "ensembl",
  organism = "Mus musculus",
  path = tempdir(),
  release = release_version
)

## import gtf
gtf <- read_gtf(gtf_file)
head(gtf)

## explore mane/appris
gtf$tag %>% unique()

## prepare tx2gene file
t2g <- gtf %>%
  dplyr::select(gene_id, transcript_id) %>%
  filter(!is.na(gene_id)) %>%
  filter(!is.na(transcript_id)) %>%
  distinct()

write_csv(t2g, paste0("./GENOMEDIR/Mmus_ensembl_v", release_version,"_tx2gene.csv"))

## which biotypes are present?
gtf %>%
  filter(type == "gene") %>%
  group_by(gene_biotype) %>%
  dplyr::summarise(n = n()) %>%
  arrange(desc(n))

```

### TSL support

- Take TSL1 and TSL2 unless the top scorers are absent
- Otherwise, take the next best scoring TSL

```{r}
## Numericate tsl column
gtf_tsl_fix <- gtf %>%
  mutate(transcript_support_level = str_replace_all(transcript_support_level, "\\s.+", "")) %>%
  mutate(transcript_support_level = if_else(transcript_support_level == "NA", NA_character_, transcript_support_level)) %>%
  mutate(transcript_support_level = as.double(transcript_support_level))

gtf_tsl_fix$transcript_support_level %>% unique()

gtf_tsl_stat <- gtf_tsl_fix %>%
  group_by(gene_id, gene_name, transcript_support_level) %>%
  summarise(
    n = n(),
  )

gtf <- gtf_tsl_fix


# ## Get minimum tsl score for each gene
# min_tsl_df <- gtf_tsl_fix %>%
#   group_by(gene_id) %>%
#   summarise(min_tsl = min(transcript_support_level, na.rm = TRUE))
# 
# ## classify genes into ideal (tsl1 or 2), take best (tsl3-5), take all (NA)
# genes_ideal_tsl <- min_tsl_df %>%
#   filter(min_tsl %in% c(1, 2)) %>% 
#   pull(gene_id)
# 
# genes_takebest_tsl <- min_tsl_df %>%
#   filter(min_tsl %in% c(3, 4, 5)) %>%
#   pull(gene_id)
# 
# 
# gene_takeall_tsl
# 
# 
# 
# 
# 
# ##
# pcnc_constitutive_list[["ENSG00000151005"]]

```

### Create list

```{r}
## filter for protein coding and ncRNA
gtf_pcnc <- gtf %>% 
  filter(gene_biotype %in% c("protein_coding", "lncRNA"))

## how many genes have both +/- strand attributes?
strand_test_df <- gtf_pcnc %>%
  filter(type == "exon") %>%
  group_by(gene_id) %>%
  summarise(strand = unique(strand) %>% sort() %>% paste0(collapse = ""))

## create a list for each gene
pcnc_list <- gtf_pcnc %>%
  split(gtf_pcnc$gene_id)

## create a filtered list of exons
pcnc_exon_list <- pcnc_list %>%
  future_imap(~{
    exon_intervals <- .x %>% filter(type == "exon")
    isoform_count <- unique(exon_intervals$transcript_id) %>% length()
    strand <- unique(exon_intervals$strand) %>% unique()
    min_transcript_support_level <- min(exon_intervals$transcript_support_level, na.rm = TRUE)
    
    output <- list(
      gene_id = .y,
      gene_name = unique(exon_intervals$gene_name),
      gene_biotype = unique(exon_intervals$gene_biotype),
      exon_intervals = exon_intervals,
      isoform_count = isoform_count,
      strand = strand,
      min_transcript_support_level = min_transcript_support_level
    )
    
    return(output)
  }, .progress = TRUE)

## save the list 
qsave(pcnc_exon_list, paste0("./data/RNA-structure/Mmus/Mmus_ensembl_v",
                             release_version,
                             "_proteincoding_noncoding_exon_list.qs")
)

## distribution of transcript isoform count per gene
pcnc_exon_list <- qread("./data/RNA-structure/Mmus/Mmus_ensembl_v113_proteincoding_noncoding_exon_list.qs")

isoform_count_df <- pcnc_exon_list %>%
  future_imap_dfr(~{
    tibble(
      gene_name = pluck(.x, "gene_name"),
      gene_biotype = pluck(.x, "gene_biotype"),
      isoform_count = pluck(.x, "isoform_count")
      )
  }, .id = "gene_id")

if(!dir.exists("./data/RNA-structure/Mmus/plots")){
  dir.create("./data/RNA-structure/Mmus/plots")
}

isoform_count_df %>%
  ggplot(aes(x = isoform_count)) + 
  geom_histogram(bins = 50) + 
  scale_y_log10() + 
  labs(title = "Number of tx isoforms per gene") + 
  facet_wrap(~ gene_biotype) + 
  theme_bw(base_size = 15)

ggsave("./data/RNA-structure/Mmus/plots/plot_tx-isoform-count-per-gene.pdf")

```


### Find constitutive intervals for multi-isoform transcripts

#### set up bedtools environment

* My conda env has bedtools 2.31.1, whereas bedtoolsr is built for 2.30.0 - just in case there are syntax errors...

```{r}
library(bedtoolsr)

conda_bedtools_bin <- "/opt/miniconda3/envs/bedtools/bin" # find this directory
Sys.setenv(PATH = paste(Sys.getenv("PATH"), conda_bedtools_bin, sep = .Platform$path.sep))
Sys.getenv("PATH")
Sys.which("bedtools")
```

#### Add TSL support

```{r}
##
pcnc_exon_list <- qread("./data/RNA-structure/Mmus/Mmus_ensembl_v113_proteincoding_noncoding_exon_list.qs") 

## add tsl support (1 and 2, or the best score) information
pcnc_exon_list <- pcnc_exon_list %>%
  map(~{
    if (.x$min_transcript_support_level <= 2) {
      .x[["tsl_filtered_transcript_ids"]] <- .x[["exon_intervals"]] %>%
        filter(transcript_support_level <= 2) %>%
        pull(transcript_id) %>% 
        unique()
      .x[["tsl_isoform_count"]] <- length(.x[["tsl_filtered_transcript_ids"]])
    } else if (.x$min_transcript_support_level > 2 & .x$min_transcript_support_level <= 5) {
      .x[["tsl_filtered_transcript_ids"]] <- .x[["exon_intervals"]] %>%
        filter(transcript_support_level == .x$min_transcript_support_level) %>%
        pull(transcript_id) %>% 
        unique()
      .x[["tsl_isoform_count"]] <- length(.x[["tsl_filtered_transcript_ids"]])
    } else {
      .x[["tsl_filtered_transcript_ids"]] <- .x[["exon_intervals"]] %>%
        pull(transcript_id) %>% 
        unique()
      .x[["tsl_isoform_count"]] <- length(.x[["tsl_filtered_transcript_ids"]])
    }
    
    return(.x)
  }, .progress = TRUE)

qsave(pcnc_exon_list, paste0("./data/RNA-structure/Mmus/Mmus_ensembl_v",
                             release_version,
                             "_tsl_proteincoding_noncoding_exon_list.qs")
)

pcnc_exon_list[[1000]]
```

#### find constitutive regions and create a list

```{r}
##
pcnc_exon_list <- qread("./data/RNA-structure/Mmus/Mmus_ensembl_v113_tsl_proteincoding_noncoding_exon_list.qs")

## find constitutive regions and create a list (for all isoforms)
pcnc_constitutive_list <- pcnc_exon_list %>%
  map(~{
    if(.x$isoform_count > 1){
      ## multiinter output
      multiinter_output <- pluck(.x, "exon_intervals") %>%
        split(.$transcript_id) %>%
        map(valr::bed_sort) %>%
        bedtoolsr::bt.multiinter(header = TRUE, names = paste(names(.), collapse = " ")) %>% 
        as_tibble() %>%
        janitor::row_to_names(row_number = 1) %>%
        mutate(across(c(start, end, num), as.numeric)) %>%
        mutate(strand = pluck(.x, "strand")) %>%
        mutate(width = end - start) %>%
        dplyr::select(chrom, start, end, width, strand, everything())
      
      .x$multiinter_output <- multiinter_output
      
      ## constitutive intervals
      constitutive_intervals <- multiinter_output %>%
        filter(num == pluck(.x, "isoform_count")) %>%
        dplyr::select(chrom, start, end, width, strand)
      
      .x$constitutive_intervals <- constitutive_intervals
      
      ## does constitutive intervals exist
      .x$constitutive_exist <- ifelse(nrow(constitutive_intervals) > 0, TRUE, FALSE)
      
      ## length of constitutive intervals
      .x$constitutive_length <- ifelse(
        nrow(constitutive_intervals) == 0,
        0,
        sum(constitutive_intervals$width)
      )
      
      return(.x)
      
    } else if(.x$isoform_count == 1){
      .x$multiinter_output <- pluck(.x, "exon_intervals") %>%
        mutate(num = 1) %>%
        dplyr::select(chrom, start, end, width, strand, num, "list" = transcript_id)
      
      .x$constitutive_intervals <- pluck(.x, "exon_intervals") %>%
        dplyr::select(chrom, start, end, width, strand)
      
      .x$constitutive_exist <- TRUE
      
      .x$constitutive_length <- pluck(.x, "exon_intervals") %>%
        pull(width) %>% sum()
      
      return(.x)
      
    }
  }, .progress = TRUE)

## save output
qsave(pcnc_constitutive_list, paste0("./data/RNA-structure/Mmus/Mmus_ensembl_v",
                                     release_version,
                                     "_proteincoding_noncoding_constitutive-region_list.qs")
)

## find constitutive regions and create a list (for all tsl supported isoforms)
tsl_pcnc_constitutive_list <- pcnc_exon_list %>%
  map(~{
    if(.x$tsl_isoform_count > 1){
      ## multiinter output
      multiinter_output <- pluck(.x, "exon_intervals") %>%
        dplyr::filter(transcript_id %in% .x$tsl_filtered_transcript_ids) %>%
        split(.$transcript_id) %>%
        map(valr::bed_sort) %>%
        bedtoolsr::bt.multiinter(header = TRUE, names = paste(names(.), collapse = " ")) %>% 
        as_tibble() %>%
        janitor::row_to_names(row_number = 1) %>%
        mutate(across(c(start, end, num), as.numeric)) %>%
        mutate(strand = pluck(.x, "strand")) %>%
        mutate(width = end - start) %>%
        dplyr::select(chrom, start, end, width, strand, everything())
      
      .x$multiinter_output <- multiinter_output
      
      ## constitutive intervals
      constitutive_intervals <- multiinter_output %>%
        filter(num == pluck(.x, "tsl_isoform_count")) %>%
        dplyr::select(chrom, start, end, width, strand)
      
      .x$constitutive_intervals <- constitutive_intervals
      
      ## does constitutive intervals exist
      .x$constitutive_exist <- ifelse(nrow(constitutive_intervals) > 0, TRUE, FALSE)
      
      ## length of constitutive intervals
      .x$constitutive_length <- ifelse(
        nrow(constitutive_intervals) == 0,
        0,
        sum(constitutive_intervals$width)
      )
      
      return(.x)
      
    } else if(.x$tsl_isoform_count == 1){
      .x$multiinter_output <- pluck(.x, "exon_intervals") %>%
        dplyr::filter(transcript_id %in% .x$tsl_filtered_transcript_ids) %>%
        mutate(num = 1) %>%
        dplyr::select(chrom, start, end, width, strand, num, "list" = transcript_id)
      
      .x$constitutive_intervals <- pluck(.x, "exon_intervals") %>%
        dplyr::filter(transcript_id %in% .x$tsl_filtered_transcript_ids) %>%
        dplyr::select(chrom, start, end, width, strand)
      
      .x$constitutive_exist <- TRUE
      
      .x$constitutive_length <- pluck(.x, "exon_intervals") %>%
        dplyr::filter(transcript_id %in% .x$tsl_filtered_transcript_ids) %>%
        pull(width) %>% sum()
      
      return(.x)
      
    }
  }, .progress = TRUE)


## save output
qsave(tsl_pcnc_constitutive_list, paste0("./data/RNA-structure/Mmus/Mmus_ensembl_v",
                                     release_version,
                                     "_tsl_proteincoding_noncoding_constitutive-region_list.qs")
)


```


## Extract RNA sequences 

### Download FASTA file

```{r}
release_version <- 113

fasta_file <- biomartr::getGenome(
  db = "ensembl",
  organism = "Mus musculus",
  release = release_version,
  gunzip = TRUE,
  path = tempdir(),
)

```

### Get transcript sequences of constitutive regions

#### Full constitutive (all isoforms)

```{r}
## constitutive list
pcnc_constitutive_list <- qread("./data/RNA-structure/Mmus/Mmus_ensembl_v113_proteincoding_noncoding_constitutive-region_list.qs")

## get fasta sequence, arrange order depending on the strand, and add character counts with "N"
pcnc_constitutive_seq_list <- pcnc_constitutive_list %>%
  map(~ {
    if (.x$constitutive_exist == TRUE) {
      bed_touse <- .x$constitutive_intervals %>%
        dplyr::mutate(name = str_c(chrom, start, end, strand, sep = ":")) %>%
        dplyr::select(chrom, start, end, name, "score" = width, strand)

      getfasta_output <- bedtoolsr::bt.getfasta(
        fi = fasta_file,
        bed = bed_touse,
        bedOut = TRUE,
        s = TRUE
      ) %>%
        as_tibble() %>%
        setNames(c("chrom", "start", "end", "name", "width", "strand", "seq")) %>%
        dplyr::select(-name) %>%
        dplyr::arrange(if (.x$strand == "+") start else desc(start)) %>%
        mutate(nchar_plus_N_cumsum = cumsum(width + 1))

      .x$constitutive_sequence_df <- getfasta_output
      return(.x)
    } else {
      .x$constitutive_sequence_df <- tibble(
        chrom = character(),
        start = numeric(),
        end = numeric(),
        width = numeric(),
        strand = character(),
        seq = character(),
        nchar_plus_N_cumsum = numeric()
      )
      return(.x)
    }
  }, .progress = TRUE)

## get concatenated target sequence collapsed for "N"
pcnc_constitutive_seq_list <- pcnc_constitutive_seq_list %>%
  map(~ {
    if (.x$constitutive_exist == TRUE) {
      constitutive_seq_Ncollapsed <- .x$constitutive_sequence_df %>%
        pull(seq) %>%
        paste0(collapse = "N")
      .x$constitutive_seq_Ncollapsed <- constitutive_seq_Ncollapsed
      return(.x)
    } else {
      .x$constitutive_seq_Ncollapsed <- NULL
      return(.x)
    }
  }, .progress = TRUE)

pcnc_constitutive_seq_list[[301]]

## save 
pcnc_constitutive_seq_list %>%
  qsave(paste0("./data/RNA-structure/Mmus/Mmus_ensembl_v",
               release_version,
               "_proteincoding_noncoding_constitutive-sequence_list.qs")
  )

```

#### TSL-supported isoforms

```{r}
## constitutive list
tsl_pcnc_constitutive_list <- qread("./data/RNA-structure/Mmus/Mmus_ensembl_v113_tsl_proteincoding_noncoding_constitutive-region_list.qs")

## get fasta sequence, arrange order depending on the strand, and add character counts with "N"
tsl_pcnc_constitutive_seq_list <- tsl_pcnc_constitutive_list %>%
  map(~ {
    if (.x$constitutive_exist == TRUE) {
      bed_touse <- .x$constitutive_intervals %>%
        dplyr::mutate(name = str_c(chrom, start, end, strand, sep = ":")) %>%
        dplyr::select(chrom, start, end, name, "score" = width, strand)

      getfasta_output <- bedtoolsr::bt.getfasta(
        fi = fasta_file,
        bed = bed_touse,
        bedOut = TRUE,
        s = TRUE
      ) %>%
        as_tibble() %>%
        setNames(c("chrom", "start", "end", "name", "width", "strand", "seq")) %>%
        dplyr::select(-name) %>%
        dplyr::arrange(if (.x$strand == "+") start else desc(start)) %>%
        mutate(nchar_plus_N_cumsum = cumsum(width + 1))

      .x$constitutive_sequence_df <- getfasta_output
      return(.x)
    } else {
      .x$constitutive_sequence_df <- tibble(
        chrom = character(),
        start = numeric(),
        end = numeric(),
        width = numeric(),
        strand = character(),
        seq = character(),
        nchar_plus_N_cumsum = numeric()
      )
      return(.x)
    }
  }, .progress = TRUE)

## get concatenated target sequence collapsed for "N"
tsl_pcnc_constitutive_seq_list <- tsl_pcnc_constitutive_seq_list %>%
  map(~ {
    if (.x$constitutive_exist == TRUE) {
      constitutive_seq_Ncollapsed <- .x$constitutive_sequence_df %>%
        pull(seq) %>%
        paste0(collapse = "N")
      .x$constitutive_seq_Ncollapsed <- constitutive_seq_Ncollapsed
      return(.x)
    } else {
      .x$constitutive_seq_Ncollapsed <- NULL
      return(.x)
    }
  }, .progress = TRUE)

tsl_pcnc_constitutive_seq_list[[301]]

## save 
tsl_pcnc_constitutive_seq_list %>%
  qsave(paste0("./data/RNA-structure/Mmus/Mmus_ensembl_v",
               release_version,
               "_tsl_proteincoding_noncoding_constitutive-sequence_list.qs")
  )

```



## Exploring the constitutive region dataset

### Summary of isoform constitutive mappings

```{r}
# - - - - - Full isoforms
pcnc_constitutive_seq_list <- 
  qread(paste0("./data/RNA-structure/Mmus/Mmus_ensembl_v",
               release_version,
               "_proteincoding_noncoding_constitutive-sequence_list.qs")
  )

## Summary dataframe
pcnc_constitutive_summary_df <- pcnc_constitutive_seq_list %>%
  imap_dfr(~{
    tibble(
      gene_id = .x$gene_id,
      gene_name = .x$gene_name,
      gene_biotype = .x$gene_biotype,
      strand = .x$strand,
      isoform_count = .x$isoform_count,
      constitutive_exist = .x$constitutive_exist,
      constitutive_length = .x$constitutive_length,
      constitutive_seq_Ncollapsed = ifelse(is.null(.x$constitutive_seq_Ncollapsed), 
                                           "NA", 
                                           as.character(.x$constitutive_seq_Ncollapsed))
    )
  }, .progress = TRUE) %>%
  mutate(constitutive_exist_over_500 = if_else(constitutive_length > 500, TRUE, FALSE)) %>%
  mutate(constitutive_exist_over_2k = if_else(constitutive_length > 2000, TRUE, FALSE))

write_csv(pcnc_constitutive_summary_df, paste0("./data/RNA-structure/Mmus/Mmus_ensembl_v",
                                               release_version,
                                               "_proteincoding_noncoding_constitutive-sequence_summary.csv")
)

## Summary plots
pcnc_constitutive_summary_df %>%
  ggplot(aes(x = gene_biotype, fill = constitutive_exist_over_500)) +
  geom_bar(position = position_stack()) + 
  geom_text(stat = "count", aes(label = after_stat(count)), position = position_stack(), vjust = -0.5) + 
  labs(title = "Constitutive sequence over 500nt exist?") + 
  theme_bw(base_size = 15) 

ggsave("./data/RNA-structure/Mmus/plots/plot_constitutive-length-over-500nt.pdf")

pcnc_constitutive_summary_df %>%
  ggplot(aes(x = gene_biotype, fill = constitutive_exist_over_2k)) +
  geom_bar(position = position_stack()) + 
  geom_text(stat = "count", aes(label = after_stat(count)), position = position_stack(), vjust = -0.5) + 
  labs(title = "Constitutive sequence over 2000nt exist?") + 
  theme_bw(base_size = 15) 

ggsave("./data/RNA-structure/Mmus/plots/plot_constitutive-length-over-2000nt.pdf")

pcnc_constitutive_summary_df %>%
  ggplot(aes(x = constitutive_length)) +
  geom_histogram() + 
  scale_x_log10() +
  facet_wrap(~ constitutive_exist + gene_biotype) +
  labs(title = "Have constitutive sequence?") + 
  theme_bw(base_size = 15)

ggsave("./data/RNA-structure/Mmus/plots/plot_constitutive-length-histogram.pdf")

# - - - - - TSL-supported isoforms
tsl_pcnc_constitutive_seq_list <- qread("./data/RNA-structure/Mmus/Mmus_ensembl_v113_tsl_proteincoding_noncoding_constitutive-sequence_list.qs")

## Summary dataframe
tsl_pcnc_constitutive_summary_df <- tsl_pcnc_constitutive_seq_list %>%
  imap_dfr(~{
    tibble(
      gene_id = .x$gene_id,
      gene_name = .x$gene_name,
      gene_biotype = .x$gene_biotype,
      strand = .x$strand,
      isoform_count = .x$tsl_isoform_count,
      constitutive_exist = .x$constitutive_exist,
      constitutive_length = .x$constitutive_length,
      constitutive_seq_Ncollapsed = ifelse(is.null(.x$constitutive_seq_Ncollapsed), 
                                           "NA", 
                                           as.character(.x$constitutive_seq_Ncollapsed))
    )
  }, .progress = TRUE) %>%
  mutate(constitutive_exist_over_500 = if_else(constitutive_length > 500, TRUE, FALSE)) %>%
  mutate(constitutive_exist_over_2k = if_else(constitutive_length > 2000, TRUE, FALSE))

write_csv(tsl_pcnc_constitutive_summary_df, paste0("./data/RNA-structure/Mmus/Mmus_ensembl_v",
                                               release_version,
                                               "_tsl_proteincoding_noncoding_constitutive-sequence_summary.csv")
)

## Summary plots
tsl_pcnc_constitutive_summary_df %>%
  ggplot(aes(x = gene_biotype, fill = constitutive_exist_over_500)) +
  geom_bar(position = position_stack()) + 
  geom_text(stat = "count", aes(label = after_stat(count)), position = position_stack(), vjust = -0.5) + 
  labs(title = "Constitutive sequence over 500nt exist?") + 
  theme_bw(base_size = 15) 

ggsave("./data/RNA-structure/Mmus/plots/plot_tsl_constitutive-length-over-500nt.pdf")

tsl_pcnc_constitutive_summary_df %>%
  ggplot(aes(x = gene_biotype, fill = constitutive_exist_over_2k)) +
  geom_bar(position = position_stack()) + 
  geom_text(stat = "count", aes(label = after_stat(count)), position = position_stack(), vjust = -0.5) + 
  labs(title = "Constitutive sequence over 2000nt exist?") + 
  theme_bw(base_size = 15) 

ggsave("./data/RNA-structure/Mmus/plots/plot_tsl_constitutive-length-over-2000nt.pdf")

tsl_pcnc_constitutive_summary_df %>%
  ggplot(aes(x = constitutive_length)) +
  geom_histogram() + 
  scale_x_log10() +
  facet_wrap(~ constitutive_exist + gene_biotype) +
  labs(title = "Have constitutive sequence?") + 
  theme_bw(base_size = 15)

ggsave("./data/RNA-structure/Mmus/plots/plot_tsl_constitutive-length-histogram.pdf")


```

### Plotting transcript architecture

```{r}
pcnc_constitutive_seq_list <- qread("./data/RNA-structure/Mmus/Mmus_ensembl_v113_proteincoding_noncoding_constitutive-sequence_list.qs")

tsl_pcnc_constitutive_seq_list <- qread("./data/RNA-structure/Mmus/Mmus_ensembl_v113_tsl_proteincoding_noncoding_constitutive-sequence_list.qs")

## helper function
plot_isoforms <- function(pcnc_seq, gene_name, colour_tsl_score = TRUE){
  list_element <- pcnc_seq %>% 
    purrr::keep(~.x[["gene_name"]] %in% gene_name) %>% pluck(1)
  
  if(colour_tsl_score){
  list_element$exon_intervals %>% ggplot() +
      geom_segment(
        aes(x = min(list_element$exon_intervals$start), 
            xend = max(list_element$exon_intervals$end), 
            y = reorder(transcript_id, -transcript_support_level), 
            yend = reorder(transcript_id, -transcript_support_level)),
        linewidth = 0.5, colour = "gray80"
      ) +
      geom_segment(
        data = list_element$exon_intervals,
        aes(x = start, xend = end, y = reorder(transcript_id, -transcript_support_level), yend = reorder(transcript_id, -transcript_support_level), colour = as.factor(transcript_support_level)),
        linewidth = 3, 
        inherit.aes = FALSE
      ) +
      geom_segment(
        data = list_element$constitutive_intervals, 
        aes(x = start, xend = end, y = "CONSTITUTIVE", yend = "CONSTITUTIVE"),
        linewidth = 7, colour = "coral", 
        inherit.aes = FALSE
      ) + 
      colorspace::scale_colour_discrete_qualitative(palette = "Dark 3", na.value = "gray40") +
      labs(
        title = paste0(
          list_element$gene_id, " ", 
          list_element$gene_name, " ", 
          list_element$strand, "ve strand ",
          paste0("(constitutive ", list_element$constitutive_length, "nt)")
        ),
        x = "Chromosomal coordinate",
        y = "Transcript ID",
        colour = "TSL score"
      ) + 
      theme_bw()
  } else {
    list_element$exon_intervals %>% ggplot() +
      geom_segment(
        aes(x = min(list_element$exon_intervals$start), 
            xend = max(list_element$exon_intervals$end), 
            y = reorder(transcript_id, -transcript_support_level), 
            yend = reorder(transcript_id, -transcript_support_level)),
        linewidth = 0.5, colour = "gray80"
      ) +
      geom_segment(
        data = list_element$exon_intervals,
        aes(x = start, xend = end, y = reorder(transcript_id, -transcript_support_level), yend = reorder(transcript_id, -transcript_support_level)),
        colour = "gray40",
        linewidth = 3, 
        inherit.aes = FALSE
      ) +
      geom_segment(
        data = list_element$constitutive_intervals, 
        aes(x = start, xend = end, y = "CONSTITUTIVE", yend = "CONSTITUTIVE"),
        linewidth = 7, colour = "coral", 
        inherit.aes = FALSE
      ) + 
      labs(
        title = paste0(
          list_element$gene_id, " ", 
          list_element$gene_name, " ", 
          list_element$strand, "ve strand ",
          paste0("(constitutive ", list_element$constitutive_length, "nt)")
        ),
        x = "Chromosomal coordinate",
        y = "Transcript ID",
        colour = "TSL score"
      ) + 
      theme_bw()
  }
}

plot_isoforms(tsl_pcnc_constitutive_seq_list, "Csf1", colour_tsl_score = TRUE)
plot_isoforms(tsl_pcnc_constitutive_seq_list, "Cplx2", colour_tsl_score = TRUE)

plot_isoforms(tsl_pcnc_constitutive_seq_list, "Ccl2", colour_tsl_score = FALSE)
ggsave("~/Desktop/constitutive-example.png", width = 9, height = 4)


## Explore some plots
library(patchwork)

# pcnc_constitutive_seq_list %>%
#   purrr::keep(~ .x$gene_name %in% c("SYNCRIP", "PURA", "PURB", "OAS1")) %>%
#   map(plot_isoforms) %>%
#   wrap_plots()

ggsave("./data/RNA-structure/Mmus/plots/plot_constitutive_txplot-example.pdf", width = 12, height = 9)

## there's a gene without a name (NA gives == a error) ENST00000456270




tsl_pcnc_constitutive_seq_list[["ENSG00000185129"]]

x$exon_intervals %>% View()
  
  
```

### Retrieving constitutive sequence of a gene 

```{r}
pcnc_constitutive_seq_list <- 
  qread(paste0("./data/RNA-structure/Mmus/Hsap_ensembl_v",
               release_version,
               "_proteincoding_noncoding_constitutive-sequence_list.qs")
  )

gene_id <- "FBgn0025790"
output_fasta_path <- "~/Desktop.fa"

fasta_lines <- c(
  paste(
   ">", gene_id, 
   pluck(pcnc_constitutive_list, gene_id) %>% pluck("gene_name"),  
   "RNA isoform-consitutive sequence (N collapsed)" 
  ),
  pluck(pcnc_constitutive_seq_list, gene_id) %>% pluck("constitutive_seq_Ncollapsed")
)

write_lines(fasta_lines, output_fasta_path)

```

## Intersect with Gerry's list

```{r}
## What are the stats for Gerry's genes?
release_version <- 113
tsl_pcnc_constitutive_summary_df <- read_csv(
  paste0("./data/RNA-structure/Mmus/Mmus_ensembl_v",
         release_version,
         "_tsl_proteincoding_noncoding_constitutive-sequence_summary.csv")
)

gerry_gene_vector <- c("Ccl2", "Ccl3", "Ccl4", "Ccl5", "Ccl5", "Ccl7", "Csf1", "Csf2",
                       "Il1b", "Il6", "Tnf", "Il17a", "Ifna1", "Ifng", 
                       "Tlr1", "Tlr2", "Tlr3", "Tlr4", "Tlr5", "Tlr6", "Tlr7", "Tlr8", "Tlr9", "Tlr11", "Tlr12", "Tlr13",
                       "Ptprc", "Cxcr2", "Prtn3", "Csf1r", "Ly6c1", "Lyve1", "Pdpn", "Pecam1")

tibble(gene_name = gerry_gene_vector) %>%
  mutate(does_exist = gene_name %in% tsl_pcnc_constitutive_summary_df$gene_name) 

gerry_gene_df <- tsl_pcnc_constitutive_summary_df %>%
  filter(gene_name %in% gerry_gene_vector) %>%
  dplyr::select(gene_name, "constitutive_length(nt)" = constitutive_length, isoform_count) %>%
  # arrange(desc(`constitutive_length(nt)`)) %>%
  arrange(gene_name)

library(gt) 

gerry_gene_df %>%
  gt() %>%
  tab_header(title = "Isoform-constitutive RNA sequence length", subtitle = "gene list from Gerry") %>%
  fmt_number(columns = `constitutive_length(nt)`, decimals = 0) %>%
  gtsave("~/Desktop/gerry_gene_constitutive_length.png")

# library(biomaRt)
# 
# mmus_ensembl <- useEnsembl(biomart = "ensembl", dataset = "mmusculus_gene_ensembl", version = 113)
# mmus_ensembl_attr <- listAttributes(mmus_ensembl)

## MGI annotation 
go_immune_response <- read_tsv("./data/RNA-structure/Mmus/MGI/GO_term_summary_immune-response-process.txt") %>%
  pull(Symbol) %>% unique()
go_signalling <- read_tsv("./data/RNA-structure/Mmus/MGI/GO_term_summary_signalling.txt") %>%
  pull(Symbol) %>% unique()
expr_hematopoietic_system <- read_tsv("./data/RNA-structure/Mmus/MGI/MGIgeneExpressionQuery_hematopoietic-system.txt") %>%
  filter(Detected == "Yes") %>%
  pull(`Gene Symbol`) %>% unique()
expr_immune_system <- read_tsv("./data/RNA-structure/Mmus/MGI/MGIgeneExpressionQuery_immune-system.txt") %>%
  filter(Detected == "Yes") %>%
  pull(`Gene Symbol`) %>% unique()

output_df <- tsl_pcnc_constitutive_summary_df %>%
  mutate(go_immune_response = if_else(gene_name %in% go_immune_response, TRUE, FALSE)) %>%
  mutate(go_signalling = if_else(gene_name %in% go_signalling, TRUE, FALSE)) %>%
  mutate(expr_hematopoietic_system = if_else(gene_name %in% expr_hematopoietic_system, TRUE, FALSE)) %>%
  mutate(expr_immune_system = if_else(gene_name %in% expr_immune_system, TRUE, FALSE)) %>%
  filter(constitutive_length >= 2000) %>%
  # filter(
  #   go_immune_response == TRUE |
  #     go_signalling == TRUE &
  #     expr_hematopoietic_system == TRUE |
  #     expr_immune_system == TRUE
  # ) %>%
  filter(go_immune_response == TRUE | go_signalling == TRUE) %>%
  filter(expr_hematopoietic_system == TRUE | expr_immune_system == TRUE) %>%
  # filter(
  #   go_immune_response == TRUE &
  #     go_signalling == TRUE &
  #     expr_hematopoietic_system == TRUE &
  #     expr_immune_system == TRUE
  # )
  dplyr::select(
    gene_id, gene_name, constitutive_length, "immune_response_function "= go_immune_response, "signalling_function" = go_signalling, "expression_hematopoietic_system" = expr_hematopoietic_system, "expression_immune_system" = expr_immune_system
  ) %>%
  arrange(desc(constitutive_length))


output_df %>%
  gt() %>%
  gtsave("~/Desktop/mouse-long-constitutive-genes_immune-related.rtf")



```












# HOMO SAPIENS

## Environment

```{r}
library(tidyverse)
library(valr)
library(furrr)
library(qs)
library(patchwork)

plan(multisession, workers = 8)

release_version <- 113

```

## Process GTF

### Get GTF and prepare an annotated list per protein-coding and non-coding genes

* Using the latest ENSEMBL **release 113** version
* ENSEMBL GTF is 1-based (use default read_gtf() option to convert automatically to 0-based coordinates)

```{r}
## Download gtf
release_version <- 113
gtf_file <- biomartr::getGTF(
  db = "ensembl",
  organism = "Homo sapiens",
  path = tempdir(),
  release = release_version
)

## import gtf
gtf <- read_gtf(gtf_file)
head(gtf)

## explore mane/appris
gtf$tag %>% unique()


## prepare tx2gene file
t2g <- gtf %>%
  dplyr::select(gene_id, transcript_id) %>%
  filter(!is.na(gene_id)) %>%
  filter(!is.na(transcript_id)) %>%
  distinct()

write_csv(t2g, paste0("./GENOMEDIR/Hsap_ensembl_v", release_version,"_tx2gene.csv"))

## which biotypes are present?
gtf %>%
  filter(type == "gene") %>%
  group_by(gene_biotype) %>%
  dplyr::summarise(n = n()) %>%
  arrange(desc(n))

```

### TSL support

- Take TSL1 and TSL2 unless the top scorers are absent
- Otherwise, take the next best scoring TSL

```{r}
## Numericate tsl column
gtf_tsl_fix <- gtf %>%
  mutate(transcript_support_level = str_replace_all(transcript_support_level, "\\s.+", "")) %>%
  mutate(transcript_support_level = if_else(transcript_support_level == "NA", NA_character_, transcript_support_level)) %>%
  mutate(transcript_support_level = as.double(transcript_support_level))

gtf_tsl_fix$transcript_support_level %>% unique()

gtf_tsl_stat <- gtf_tsl_fix %>%
  group_by(gene_id, gene_name, transcript_support_level) %>%
  summarise(
    n = n(),
  )

gtf <- gtf_tsl_fix


# ## Get minimum tsl score for each gene
# min_tsl_df <- gtf_tsl_fix %>%
#   group_by(gene_id) %>%
#   summarise(min_tsl = min(transcript_support_level, na.rm = TRUE))
# 
# ## classify genes into ideal (tsl1 or 2), take best (tsl3-5), take all (NA)
# genes_ideal_tsl <- min_tsl_df %>%
#   filter(min_tsl %in% c(1, 2)) %>% 
#   pull(gene_id)
# 
# genes_takebest_tsl <- min_tsl_df %>%
#   filter(min_tsl %in% c(3, 4, 5)) %>%
#   pull(gene_id)
# 
# 
# gene_takeall_tsl
# 
# 
# 
# 
# 
# ##
# pcnc_constitutive_list[["ENSG00000151005"]]

```

### Create list

```{r}
## filter for protein coding and ncRNA
gtf_pcnc <- gtf %>% 
  filter(gene_biotype %in% c("protein_coding", "lncRNA"))

## how many genes have both +/- strand attributes?
strand_test_df <- gtf_pcnc %>%
  filter(type == "exon") %>%
  group_by(gene_id) %>%
  summarise(strand = unique(strand) %>% sort() %>% paste0(collapse = ""))

## create a list for each gene
pcnc_list <- gtf_pcnc %>%
  split(gtf_pcnc$gene_id)

## create a filtered list of exons
pcnc_exon_list <- pcnc_list %>%
  future_imap(~{
    exon_intervals <- .x %>% filter(type == "exon")
    isoform_count <- unique(exon_intervals$transcript_id) %>% length()
    strand <- unique(exon_intervals$strand) %>% unique()
    min_transcript_support_level <- min(exon_intervals$transcript_support_level, na.rm = TRUE)
    
    output <- list(
      gene_id = .y,
      gene_name = unique(exon_intervals$gene_name),
      gene_biotype = unique(exon_intervals$gene_biotype),
      exon_intervals = exon_intervals,
      isoform_count = isoform_count,
      strand = strand,
      min_transcript_support_level = min_transcript_support_level
    )
    
    return(output)
  }, .progress = TRUE)

## save the list 
qsave(pcnc_exon_list, paste0("./data/RNA-structure/Hsap/Hsap_ensembl_v",
                             release_version,
                             "_proteincoding_noncoding_exon_list.qs")
)

## distribution of transcript isoform count per gene
pcnc_exon_list <- qread("./data/RNA-structure/Hsap/Hsap_ensembl_v113_proteincoding_noncoding_exon_list.qs")

isoform_count_df <- pcnc_exon_list %>%
  future_imap_dfr(~{
    tibble(
      gene_name = pluck(.x, "gene_name"),
      gene_biotype = pluck(.x, "gene_biotype"),
      isoform_count = pluck(.x, "isoform_count")
      )
  }, .id = "gene_id")

if(!dir.exists("./data/RNA-structure/Hsap/plots")){
  dir.create("./data/RNA-structure/Hsap/plots")
}

isoform_count_df %>%
  ggplot(aes(x = isoform_count)) + 
  geom_histogram(bins = 50) + 
  scale_y_log10() + 
  labs(title = "Number of tx isoforms per gene") + 
  facet_wrap(~ gene_biotype) + 
  theme_bw(base_size = 15)

ggsave("./data/RNA-structure/Hsap/plots/plot_tx-isoform-count-per-gene.pdf")

```


### Find constitutive intervals for multi-isoform transcripts

#### set up bedtools environment

* My conda env has bedtools 2.31.1, whereas bedtoolsr is built for 2.30.0 - just in case there are syntax errors...

```{r}
library(bedtoolsr)

conda_bedtools_bin <- "/opt/miniconda3/envs/bedtools/bin" # find this directory
Sys.setenv(PATH = paste(Sys.getenv("PATH"), conda_bedtools_bin, sep = .Platform$path.sep))
Sys.getenv("PATH")
Sys.which("bedtools")
```

#### Add TSL support

```{r}
##
pcnc_exon_list <- qread("./data/RNA-structure/Hsap/Hsap_ensembl_v113_proteincoding_noncoding_exon_list.qs") 

## add tsl support (1 and 2, or the best score) information
pcnc_exon_list <- pcnc_exon_list %>%
  map(~{
    if (.x$min_transcript_support_level <= 2) {
      .x[["tsl_filtered_transcript_ids"]] <- .x[["exon_intervals"]] %>%
        filter(transcript_support_level <= 2) %>%
        pull(transcript_id) %>% 
        unique()
      .x[["tsl_isoform_count"]] <- length(.x[["tsl_filtered_transcript_ids"]])
    } else if (.x$min_transcript_support_level > 2 & .x$min_transcript_support_level <= 5) {
      .x[["tsl_filtered_transcript_ids"]] <- .x[["exon_intervals"]] %>%
        filter(transcript_support_level == .x$min_transcript_support_level) %>%
        pull(transcript_id) %>% 
        unique()
      .x[["tsl_isoform_count"]] <- length(.x[["tsl_filtered_transcript_ids"]])
    } else {
      .x[["tsl_filtered_transcript_ids"]] <- .x[["exon_intervals"]] %>%
        pull(transcript_id) %>% 
        unique()
      .x[["tsl_isoform_count"]] <- length(.x[["tsl_filtered_transcript_ids"]])
    }
    
    return(.x)
  }, .progress = TRUE)

qsave(pcnc_exon_list, paste0("./data/RNA-structure/Hsap/Hsap_ensembl_v",
                             release_version,
                             "_tsl_proteincoding_noncoding_exon_list.qs")
)

pcnc_exon_list[[1000]]
```

#### find constitutive regions and create a list

```{r}
##
pcnc_exon_list <- qread("./data/RNA-structure/Hsap/Hsap_ensembl_v113_tsl_proteincoding_noncoding_exon_list.qs")

## find constitutive regions and create a list (for all isoforms)
pcnc_constitutive_list <- pcnc_exon_list %>%
  map(~{
    if(.x$isoform_count > 1){
      ## multiinter output
      multiinter_output <- pluck(.x, "exon_intervals") %>%
        split(.$transcript_id) %>%
        map(valr::bed_sort) %>%
        bedtoolsr::bt.multiinter(header = TRUE, names = paste(names(.), collapse = " ")) %>% 
        as_tibble() %>%
        janitor::row_to_names(row_number = 1) %>%
        mutate(across(c(start, end, num), as.numeric)) %>%
        mutate(strand = pluck(.x, "strand")) %>%
        mutate(width = end - start) %>%
        dplyr::select(chrom, start, end, width, strand, everything())
      
      .x$multiinter_output <- multiinter_output
      
      ## constitutive intervals
      constitutive_intervals <- multiinter_output %>%
        filter(num == pluck(.x, "isoform_count")) %>%
        dplyr::select(chrom, start, end, width, strand)
      
      .x$constitutive_intervals <- constitutive_intervals
      
      ## does constitutive intervals exist
      .x$constitutive_exist <- ifelse(nrow(constitutive_intervals) > 0, TRUE, FALSE)
      
      ## length of constitutive intervals
      .x$constitutive_length <- ifelse(
        nrow(constitutive_intervals) == 0,
        0,
        sum(constitutive_intervals$width)
      )
      
      return(.x)
      
    } else if(.x$isoform_count == 1){
      .x$multiinter_output <- pluck(.x, "exon_intervals") %>%
        mutate(num = 1) %>%
        dplyr::select(chrom, start, end, width, strand, num, "list" = transcript_id)
      
      .x$constitutive_intervals <- pluck(.x, "exon_intervals") %>%
        dplyr::select(chrom, start, end, width, strand)
      
      .x$constitutive_exist <- TRUE
      
      .x$constitutive_length <- pluck(.x, "exon_intervals") %>%
        pull(width) %>% sum()
      
      return(.x)
      
    }
  }, .progress = TRUE)

## save output
qsave(pcnc_constitutive_list, paste0("./data/RNA-structure/Hsap/Hsap_ensembl_v",
                                     release_version,
                                     "_proteincoding_noncoding_constitutive-region_list.qs")
)

## find constitutive regions and create a list (for all tsl supported isoforms)
tsl_pcnc_constitutive_list <- pcnc_exon_list %>%
  map(~{
    if(.x$tsl_isoform_count > 1){
      ## multiinter output
      multiinter_output <- pluck(.x, "exon_intervals") %>%
        dplyr::filter(transcript_id %in% .x$tsl_filtered_transcript_ids) %>%
        split(.$transcript_id) %>%
        map(valr::bed_sort) %>%
        bedtoolsr::bt.multiinter(header = TRUE, names = paste(names(.), collapse = " ")) %>% 
        as_tibble() %>%
        janitor::row_to_names(row_number = 1) %>%
        mutate(across(c(start, end, num), as.numeric)) %>%
        mutate(strand = pluck(.x, "strand")) %>%
        mutate(width = end - start) %>%
        dplyr::select(chrom, start, end, width, strand, everything())
      
      .x$multiinter_output <- multiinter_output
      
      ## constitutive intervals
      constitutive_intervals <- multiinter_output %>%
        filter(num == pluck(.x, "tsl_isoform_count")) %>%
        dplyr::select(chrom, start, end, width, strand)
      
      .x$constitutive_intervals <- constitutive_intervals
      
      ## does constitutive intervals exist
      .x$constitutive_exist <- ifelse(nrow(constitutive_intervals) > 0, TRUE, FALSE)
      
      ## length of constitutive intervals
      .x$constitutive_length <- ifelse(
        nrow(constitutive_intervals) == 0,
        0,
        sum(constitutive_intervals$width)
      )
      
      return(.x)
      
    } else if(.x$tsl_isoform_count == 1){
      .x$multiinter_output <- pluck(.x, "exon_intervals") %>%
        dplyr::filter(transcript_id %in% .x$tsl_filtered_transcript_ids) %>%
        mutate(num = 1) %>%
        dplyr::select(chrom, start, end, width, strand, num, "list" = transcript_id)
      
      .x$constitutive_intervals <- pluck(.x, "exon_intervals") %>%
        dplyr::filter(transcript_id %in% .x$tsl_filtered_transcript_ids) %>%
        dplyr::select(chrom, start, end, width, strand)
      
      .x$constitutive_exist <- TRUE
      
      .x$constitutive_length <- pluck(.x, "exon_intervals") %>%
        dplyr::filter(transcript_id %in% .x$tsl_filtered_transcript_ids) %>%
        pull(width) %>% sum()
      
      return(.x)
      
    }
  }, .progress = TRUE)


## save output
qsave(tsl_pcnc_constitutive_list, paste0("./data/RNA-structure/Hsap/Hsap_ensembl_v",
                                     release_version,
                                     "_tsl_proteincoding_noncoding_constitutive-region_list.qs")
)


```


## Extract RNA sequences 

### Download FASTA file

```{r}
release_version <- 113

fasta_file <- biomartr::getGenome(
  db = "ensembl",
  organism = "Homo sapiens",
  release = release_version,
  gunzip = TRUE,
  path = tempdir(),
)

```

### Get transcript sequences of constitutive regions

#### Full constitutive (all isoforms)

```{r}
## constitutive list
pcnc_constitutive_list <- qread("./data/RNA-structure/Hsap/Hsap_ensembl_v113_proteincoding_noncoding_constitutive-region_list.qs")

## get fasta sequence, arrange order depending on the strand, and add character counts with "N"
pcnc_constitutive_seq_list <- pcnc_constitutive_list %>%
  map(~ {
    if (.x$constitutive_exist == TRUE) {
      bed_touse <- .x$constitutive_intervals %>%
        dplyr::mutate(name = str_c(chrom, start, end, strand, sep = ":")) %>%
        dplyr::select(chrom, start, end, name, "score" = width, strand)

      getfasta_output <- bedtoolsr::bt.getfasta(
        fi = fasta_file,
        bed = bed_touse,
        bedOut = TRUE,
        s = TRUE
      ) %>%
        as_tibble() %>%
        setNames(c("chrom", "start", "end", "name", "width", "strand", "seq")) %>%
        dplyr::select(-name) %>%
        dplyr::arrange(if (.x$strand == "+") start else desc(start)) %>%
        mutate(nchar_plus_N_cumsum = cumsum(width + 1))

      .x$constitutive_sequence_df <- getfasta_output
      return(.x)
    } else {
      .x$constitutive_sequence_df <- tibble(
        chrom = character(),
        start = numeric(),
        end = numeric(),
        width = numeric(),
        strand = character(),
        seq = character(),
        nchar_plus_N_cumsum = numeric()
      )
      return(.x)
    }
  }, .progress = TRUE)

## get concatenated target sequence collapsed for "N"
pcnc_constitutive_seq_list <- pcnc_constitutive_seq_list %>%
  map(~ {
    if (.x$constitutive_exist == TRUE) {
      constitutive_seq_Ncollapsed <- .x$constitutive_sequence_df %>%
        pull(seq) %>%
        paste0(collapse = "N")
      .x$constitutive_seq_Ncollapsed <- constitutive_seq_Ncollapsed
      return(.x)
    } else {
      .x$constitutive_seq_Ncollapsed <- NULL
      return(.x)
    }
  }, .progress = TRUE)

pcnc_constitutive_seq_list[[301]]

## save 
pcnc_constitutive_seq_list %>%
  qsave(paste0("./data/RNA-structure/Hsap/Hsap_ensembl_v",
               release_version,
               "_proteincoding_noncoding_constitutive-sequence_list.qs")
  )

```

#### TSL-supported isoforms

```{r}
## constitutive list
tsl_pcnc_constitutive_list <- qread("./data/RNA-structure/Hsap/Hsap_ensembl_v113_tsl_proteincoding_noncoding_constitutive-region_list.qs")

## get fasta sequence, arrange order depending on the strand, and add character counts with "N"
tsl_pcnc_constitutive_seq_list <- tsl_pcnc_constitutive_list %>%
  map(~ {
    if (.x$constitutive_exist == TRUE) {
      bed_touse <- .x$constitutive_intervals %>%
        dplyr::mutate(name = str_c(chrom, start, end, strand, sep = ":")) %>%
        dplyr::select(chrom, start, end, name, "score" = width, strand)

      getfasta_output <- bedtoolsr::bt.getfasta(
        fi = fasta_file,
        bed = bed_touse,
        bedOut = TRUE,
        s = TRUE
      ) %>%
        as_tibble() %>%
        setNames(c("chrom", "start", "end", "name", "width", "strand", "seq")) %>%
        dplyr::select(-name) %>%
        dplyr::arrange(if (.x$strand == "+") start else desc(start)) %>%
        mutate(nchar_plus_N_cumsum = cumsum(width + 1))

      .x$constitutive_sequence_df <- getfasta_output
      return(.x)
    } else {
      .x$constitutive_sequence_df <- tibble(
        chrom = character(),
        start = numeric(),
        end = numeric(),
        width = numeric(),
        strand = character(),
        seq = character(),
        nchar_plus_N_cumsum = numeric()
      )
      return(.x)
    }
  }, .progress = TRUE)

## get concatenated target sequence collapsed for "N"
tsl_pcnc_constitutive_seq_list <- tsl_pcnc_constitutive_seq_list %>%
  map(~ {
    if (.x$constitutive_exist == TRUE) {
      constitutive_seq_Ncollapsed <- .x$constitutive_sequence_df %>%
        pull(seq) %>%
        paste0(collapse = "N")
      .x$constitutive_seq_Ncollapsed <- constitutive_seq_Ncollapsed
      return(.x)
    } else {
      .x$constitutive_seq_Ncollapsed <- NULL
      return(.x)
    }
  }, .progress = TRUE)

tsl_pcnc_constitutive_seq_list[[301]]

## save 
tsl_pcnc_constitutive_seq_list %>%
  qsave(paste0("./data/RNA-structure/Hsap/Hsap_ensembl_v",
               release_version,
               "_tsl_proteincoding_noncoding_constitutive-sequence_list.qs")
  )

```



## Exploring the constitutive region dataset

### Summary of isoform constitutive mappings

```{r}
# - - - - - Full isoforms
pcnc_constitutive_seq_list <- 
  qread(paste0("./data/RNA-structure/Hsap/Hsap_ensembl_v",
               release_version,
               "_proteincoding_noncoding_constitutive-sequence_list.qs")
  )

## Summary dataframe
pcnc_constitutive_summary_df <- pcnc_constitutive_seq_list %>%
  imap_dfr(~{
    tibble(
      gene_id = .x$gene_id,
      gene_name = .x$gene_name,
      gene_biotype = .x$gene_biotype,
      strand = .x$strand,
      isoform_count = .x$isoform_count,
      constitutive_exist = .x$constitutive_exist,
      constitutive_length = .x$constitutive_length,
      constitutive_seq_Ncollapsed = ifelse(is.null(.x$constitutive_seq_Ncollapsed), 
                                           "NA", 
                                           as.character(.x$constitutive_seq_Ncollapsed))
    )
  }, .progress = TRUE) %>%
  mutate(constitutive_exist_over_500 = if_else(constitutive_length > 500, TRUE, FALSE)) %>%
  mutate(constitutive_exist_over_2k = if_else(constitutive_length > 2000, TRUE, FALSE))

write_csv(pcnc_constitutive_summary_df, paste0("./data/RNA-structure/Hsap/Hsap_ensembl_v",
                                               release_version,
                                               "_proteincoding_noncoding_constitutive-sequence_summary.csv")
)

## Summary plots
pcnc_constitutive_summary_df %>%
  ggplot(aes(x = gene_biotype, fill = constitutive_exist_over_500)) +
  geom_bar(position = position_stack()) + 
  geom_text(stat = "count", aes(label = after_stat(count)), position = position_stack(), vjust = -0.5) + 
  labs(title = "Constitutive sequence over 500nt exist?") + 
  theme_bw(base_size = 15) 

ggsave("./data/RNA-structure/Hsap/plots/plot_constitutive-length-over-500nt.pdf")

pcnc_constitutive_summary_df %>%
  ggplot(aes(x = gene_biotype, fill = constitutive_exist_over_2k)) +
  geom_bar(position = position_stack()) + 
  geom_text(stat = "count", aes(label = after_stat(count)), position = position_stack(), vjust = -0.5) + 
  labs(title = "Constitutive sequence over 2000nt exist?") + 
  theme_bw(base_size = 15) 

ggsave("./data/RNA-structure/Hsap/plots/plot_constitutive-length-over-2000nt.pdf")

pcnc_constitutive_summary_df %>%
  ggplot(aes(x = constitutive_length)) +
  geom_histogram() + 
  scale_x_log10() +
  facet_wrap(~ constitutive_exist + gene_biotype) +
  labs(title = "Have constitutive sequence?") + 
  theme_bw(base_size = 15)

ggsave("./data/RNA-structure/Hsap/plots/plot_constitutive-length-histogram.pdf")

# - - - - - TSL-supported isoforms
tsl_pcnc_constitutive_seq_list <- qread("./data/RNA-structure/Hsap/Hsap_ensembl_v113_tsl_proteincoding_noncoding_constitutive-sequence_list.qs")

## Summary dataframe
tsl_pcnc_constitutive_summary_df <- tsl_pcnc_constitutive_seq_list %>%
  imap_dfr(~{
    tibble(
      gene_id = .x$gene_id,
      gene_name = .x$gene_name,
      gene_biotype = .x$gene_biotype,
      strand = .x$strand,
      isoform_count = .x$tsl_isoform_count,
      constitutive_exist = .x$constitutive_exist,
      constitutive_length = .x$constitutive_length,
      constitutive_seq_Ncollapsed = ifelse(is.null(.x$constitutive_seq_Ncollapsed), 
                                           "NA", 
                                           as.character(.x$constitutive_seq_Ncollapsed))
    )
  }, .progress = TRUE) %>%
  mutate(constitutive_exist_over_500 = if_else(constitutive_length > 500, TRUE, FALSE)) %>%
  mutate(constitutive_exist_over_2k = if_else(constitutive_length > 2000, TRUE, FALSE))

write_csv(tsl_pcnc_constitutive_summary_df, paste0("./data/RNA-structure/Hsap/Hsap_ensembl_v",
                                               release_version,
                                               "_tsl_proteincoding_noncoding_constitutive-sequence_summary.csv")
)

## Summary plots
tsl_pcnc_constitutive_summary_df %>%
  ggplot(aes(x = gene_biotype, fill = constitutive_exist_over_500)) +
  geom_bar(position = position_stack()) + 
  geom_text(stat = "count", aes(label = after_stat(count)), position = position_stack(), vjust = -0.5) + 
  labs(title = "Constitutive sequence over 500nt exist?") + 
  theme_bw(base_size = 15) 

ggsave("./data/RNA-structure/Hsap/plots/plot_tsl_constitutive-length-over-500nt.pdf")

tsl_pcnc_constitutive_summary_df %>%
  ggplot(aes(x = gene_biotype, fill = constitutive_exist_over_2k)) +
  geom_bar(position = position_stack()) + 
  geom_text(stat = "count", aes(label = after_stat(count)), position = position_stack(), vjust = -0.5) + 
  labs(title = "Constitutive sequence over 2000nt exist?") + 
  theme_bw(base_size = 15) 

ggsave("./data/RNA-structure/Hsap/plots/plot_tsl_constitutive-length-over-2000nt.pdf")

tsl_pcnc_constitutive_summary_df %>%
  ggplot(aes(x = constitutive_length)) +
  geom_histogram() + 
  scale_x_log10() +
  facet_wrap(~ constitutive_exist + gene_biotype) +
  labs(title = "Have constitutive sequence?") + 
  theme_bw(base_size = 15)

ggsave("./data/RNA-structure/Hsap/plots/plot_tsl_constitutive-length-histogram.pdf")


```

### Plotting transcript architecture

```{r}
pcnc_constitutive_seq_list <- qread("./data/RNA-structure/Hsap/Hsap_ensembl_v113_proteincoding_noncoding_constitutive-sequence_list.qs")

tsl_pcnc_constitutive_seq_list <- qread("./data/RNA-structure/Hsap/Hsap_ensembl_v113_tsl_proteincoding_noncoding_constitutive-sequence_list.qs")

## helper function
plot_isoforms <- function(pcnc_seq, gene_name, colour_tsl_score = TRUE){
  list_element <- pcnc_seq %>% 
    purrr::keep(~.x[["gene_name"]] %in% gene_name) %>% pluck(1)
  
  if(colour_tsl_score){
  list_element$exon_intervals %>% ggplot() +
      geom_segment(
        aes(x = min(list_element$exon_intervals$start), 
            xend = max(list_element$exon_intervals$end), 
            y = reorder(transcript_id, -transcript_support_level), 
            yend = reorder(transcript_id, -transcript_support_level)),
        linewidth = 0.5, colour = "gray80"
      ) +
      geom_segment(
        data = list_element$exon_intervals,
        aes(x = start, xend = end, y = reorder(transcript_id, -transcript_support_level), yend = reorder(transcript_id, -transcript_support_level), colour = as.factor(transcript_support_level)),
        linewidth = 3, 
        inherit.aes = FALSE
      ) +
      geom_segment(
        data = list_element$constitutive_intervals, 
        aes(x = start, xend = end, y = "CONSTITUTIVE", yend = "CONSTITUTIVE"),
        linewidth = 7, colour = "coral", 
        inherit.aes = FALSE
      ) + 
      colorspace::scale_colour_discrete_qualitative(palette = "Dark 3", na.value = "gray40") +
      labs(
        title = paste0(
          list_element$gene_id, " ", 
          list_element$gene_name, " ", 
          list_element$strand, "ve strand ",
          paste0("(constitutive ", list_element$constitutive_length, "nt)")
        ),
        x = "Chromosomal coordinate",
        y = "Transcript ID",
        colour = "TSL score"
      ) + 
      theme_bw()
  } else {
    list_element$exon_intervals %>% ggplot() +
      geom_segment(
        aes(x = min(list_element$exon_intervals$start), 
            xend = max(list_element$exon_intervals$end), 
            y = reorder(transcript_id, -transcript_support_level), 
            yend = reorder(transcript_id, -transcript_support_level)),
        linewidth = 0.5, colour = "gray80"
      ) +
      geom_segment(
        data = list_element$exon_intervals,
        aes(x = start, xend = end, y = reorder(transcript_id, -transcript_support_level), yend = reorder(transcript_id, -transcript_support_level)),
        colour = "gray40",
        linewidth = 3, 
        inherit.aes = FALSE
      ) +
      geom_segment(
        data = list_element$constitutive_intervals, 
        aes(x = start, xend = end, y = "CONSTITUTIVE", yend = "CONSTITUTIVE"),
        linewidth = 7, colour = "coral", 
        inherit.aes = FALSE
      ) + 
      labs(
        title = paste0(
          list_element$gene_id, " ", 
          list_element$gene_name, " ", 
          list_element$strand, "ve strand ",
          paste0("(constitutive ", list_element$constitutive_length, "nt)")
        ),
        x = "Chromosomal coordinate",
        y = "Transcript ID",
        colour = "TSL score"
      ) + 
      theme_bw()
  }
}

plot_isoforms(pcnc_constitutive_seq_list, "OAS1", colour_tsl_score = TRUE)
plot_isoforms(tsl_pcnc_constitutive_seq_list, "TMPRSS2", colour_tsl_score = TRUE)



## Explore some plots
library(patchwork)

# pcnc_constitutive_seq_list %>%
#   purrr::keep(~ .x$gene_name %in% c("SYNCRIP", "PURA", "PURB", "OAS1")) %>%
#   map(plot_isoforms) %>%
#   wrap_plots()

ggsave("./data/RNA-structure/Hsap/plots/plot_constitutive_txplot-example.pdf", width = 12, height = 9)

## there's a gene without a name (NA gives == a error) ENST00000456270




tsl_pcnc_constitutive_seq_list[["ENSG00000185129"]]

x$exon_intervals %>% View()
  
  
```

### Retrieving constitutive sequence of a gene 

```{r}
pcnc_constitutive_seq_list <- 
  qread(paste0("./data/RNA-structure/Hsap/Hsap_ensembl_v",
               release_version,
               "_proteincoding_noncoding_constitutive-sequence_list.qs")
  )

gene_id <- "FBgn0025790"
output_fasta_path <- "~/Desktop.fa"

fasta_lines <- c(
  paste(
   ">", gene_id, 
   pluck(pcnc_constitutive_list, gene_id) %>% pluck("gene_name"),  
   "RNA isoform-consitutive sequence (N collapsed)" 
  ),
  pluck(pcnc_constitutive_seq_list, gene_id) %>% pluck("constitutive_seq_Ncollapsed")
)

write_lines(fasta_lines, output_fasta_path)

```























































